Multidimensional analysis and detection of informative features in human brain white matter

### 2021-06-09 reports/manuscript/paper.tex
Fix PLOS formatting requests

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in human brain white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{1,4},
\\
\bigskip
\textbf{1} eScience Institute, University of Washington, Seattle, WA, USA
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, CA, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, USA
\\
\textbf{4} Department of Psychology, University of Washington, Seattle, WA, USA
\\
\bigskip

@@

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL and PCR-SGL also identify
the white matter tracts most important for ALS classification.
\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \cite{sarica2017corticospinal}.
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups (see supplemental Figs S1 and S2 for tract profiles of all 18 tracts in these groups).
We find that SGL and PCR-SGL select FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the tract profile
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets see supplemental Figs S3 to S8 for the full tract profile information in all tracts/datasets.

@@

In contrast to the ALS classification case, the selected $\alpha$ values
indicate high global sparsity over group sparsity, with $\alpha = $
{\whLRatio}, {\hbnLRatio}, and {\ccLRatio}, for the WH, HBN, and Cam-CAN
datasets, respectively. The model weights are distributed over many different
tracts and dMRI tissue properties (see
\cref{fig:age-results:wh-coefs3d,fig:age-results:hbn-coefs3d,fig:age-results:cc-coefs3d}
and supporing figures in \nameref{S1_Appendix}).
This demonstrates that SGL is not coerced to produce overly sparse results
when a more accurate model requires a dense selection of features.
Furthermore, inspecting the portions of bundles with larger coefficients in
\cref{fig:age-results:wh-profile,fig:age-results:hbn-profile,fig:age-results:cc-profile}
reveals that SGL selects informative regions where diffusion properties are
different between the age quintiles.

@@

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \cite{pinto2020harmonization}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.

@@

\item Diffusion MRI from a previous study of the corticospinal
tract (CST) in patients with amyotrophic lateral sclerosis
(ALS \cite{sarica2017corticospinal}), containing data from 24 ALS
patients and 24 demographically matched healthy controls. These data
were measured in a GE Discovery 750 3T MRI scanner at the Institute
of Bioimaging and Molecular Physiology in Catanzaro. Informed consent
was provided as approved by the Ethical Committee of the University
``Magna Graecia'' of Catanzaro. Voxel resolution was \num{2x2x2}
$\text{mm}^3$ and 27 non-colinear directions were measured with a
$b=1000$ $\text{s} / \text{mm}^2$. Data was preprocessed to
correct for subject motion and for eddy currents. The diffusion tensor
model \cite{basser1994mr} was fit in every voxel.
We will refer to this dataset as ALS.

\item Diffusion MRI data from a previous study of properties of
the white matter across the lifespan \cite{yeatman2014lifespan},
containing dMRI data from 76 subjects with ages 6-50. These data were
measured in a GE Discovery 750 3T MRI scanner at the Stanford Center
for Cognitive and Neurobiological Imaging. The Stanford University
IRB approved the procedures of this study. Informed consent was
obtained from each adult participant, and assent for participation
was provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}$\textrm{mm}^3$ with 96 non-colinear directions measured with a
$b=2000$ $\textrm{s} / \textrm{mm}^2$ and 30 non-colinear directions
measured with a $b=1000$ $\textrm{s} / \textrm{mm}^2$. These data
were acquired using a dual spin echo sequence, in which there is
sufficient time for eddy currents to subside between the application of
the gradients and the image acquisition, so no eddy current correction
was applied, but motion correction was applied before fitting the
diffusion tensor model \cite{basser1994mr} in every voxel using a robust
fit \cite{chang2005restore}
on the \protect$b=1000$ \protect$\textrm{s} / \textrm{mm}^2$ shell only.
We will refer to this dataset as WH.

\item Diffusion MRI data from the Healthy Brain Network pediatric mental
health study \cite{alexander2017open}, containing dMRI data from 1651 subjects
with ages 5-21. These data were measured in 3T Siemens MRI scanners at two
sites in the New York
area: Rutgers University Brain Imaging Center and the CitiGroup Cornell Brain Imaging Center.
Informed consent was obtained from each participant aged 18 or older. For
participants younger than 18, written consent was obtained from their legal
guardians and written assent was obtained from the participant. Voxel
resolution was \num{1.8x1.8x1.8}$\text{mm}^3$ with 64 non-colinear directions
measured for each of $b=1000 \text{s} / \text{mm}^2$ and $b=2000 \text{s} /
\text{mm}^2$. Preprocessing was performed using \emph{QSIPrep} 0.12.1, which
is based on \emph{Nipype} 1.5.1 \cite[RRID:SCR\_002502]{nipype1,nipype2}.

@@

\item Diffusion MRI data from the Cambridge Centre for Ageing and
Neuroscience (Cam-CAN) ``CC700'' dataset
\cite{shafto2014cambridge,taylor2017cambridge}, containing data from 640
subjects with ages 18-88. These data were measured on a 3T Siemens TIM Trio
system and written informed consent was obtained from each participant. Voxel
resolution was \num{2x2x2}$\text{mm}^3$ with 30 non-colinear directions
measured for each of $b=1000$ $\text{s} / \text{mm}^2$ and $b=2000$
$\text{s} / \text{mm}^2$. The diffusion weighted images were acquired
with a twice refocused spin-echo sequence and the same preprocessing
and reconstruction pipelines
used for the HBN dataset was applied to this data. We will refer to this
dataset as Cam-CAN.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox
(\texttt{mAFQ}, version 1.1 for WH and version 1.2 for ALS)
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Eighteen major
tracts, which are enumerated in the supplemental material, were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalanobis distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the supplemental material.
See Fig S3 of \cite{kruper2021evaluating} for a depiction of these white matter bundles.
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (see \cref{fig:methods:group-structure}).

@@

For the multisite HBN study, we use the ComBat harmonization method to robustly adjust for site effects in the tract profiles. Initially designed to correct for site effects in gene expression studies \cite{johnson2007adjusting}, ComBat employs a parametric empirical Bayes approach to adjust for batch effects and has since been applied to multi-site cortical thickness measurements \cite{fortin2018harmonization}, multi-site DTI studies \cite{fortin2017harmonization}, and functional MRI data in the Adolescent Brain Cognitive Development Study (ABCD) \cite{nielson2018detecting}. We rely on the \texttt{neurocombat\_sklearn} library \cite{neurocombat_sklearn}, to apply ComBat in our scikit-learn analysis pipeline and present bundle profile site differences in the supplemental material.

@@

@@

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in human brain white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{1,4},
\\
\bigskip
\textbf{1} eScience Institute, University of Washington, Seattle, Washington, United States of America
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, California, United States of America
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, Washington, United States of America
\\
\textbf{4} Department of Psychology, University of Washington, Seattle, Washington, United States of America
\\
\bigskip

@@

\begin{figure}[b!]
    % \vspace{0.25cm}
    % \rlap{
    %     \hspace{-0.25em}
    %     \includegraphics[width=0.98\columnwidth]{methods-brains.pdf}
    % }
    % \vspace{-8.4cm}
    % \includegraphics[width=\columnwidth]{method-quad.pdf}
    {\phantomsubcaption\label{fig:methods:tractogram}}
    {\phantomsubcaption\label{fig:methods:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:fa}}
    {\phantomsubcaption\label{fig:methods:group-structure}}
    \caption{\label{fig:methods}
    {\bf Tractometry data flow}
    \textbf{(a)} Whole brain tractography generates streamlines approximating
        the trajectories of white matter connections.
        \textbf{(b)} Tractometry classifies these streamlines into anatomical bundles.
        In this case, we show the left corticospinal tract (CSTL)
        and the left arcuate fasciculus (ARCL)
        over a mid-saggital anatomical slice.
        Tract profiling further extracts bundle profiles,
        quantifications of various diffusion metrics along the length of the
        fiber bundle. Here, we show one subject's fractional anisotropy (FA)
        profile for \textbf{(c)} the CSTL and \textbf{(d)} the ARCL.
        \textbf{(d)} the phenotypical target data and tract profile
        features can be organized into a linear model, $\hat{y} = \mathbf{X}
        \hat{\beta}$. The feature matrix $\mathbf{X}$ is color-coded
        to reveal a natural group structure: the left (orange) group
        contains $k$ features from the CSTL, the middle (green) group
        contains $k$ features from the left cingulum cingulate (CGCL),
        and the right (blue) group
        contains $k$ features from the ARCL.
        The coefficients in $\hat{\beta}$ follow the same natural grouping. Panels (a) and (b) are adapted from \url{https://figshare.com/articles/figure/example_tractography-segmentation/14485350}, and reproduced under the CC-BY license (\url{https://creativecommons.org/licenses/by/4.0/})
    }
\end{figure}

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL and PCR-SGL also identify
the white matter tracts most important for ALS classification.
\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \cite{sarica2017corticospinal}.
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups
(see \nameref{S1_Fig} and \nameref{S2_Fig} for tract profiles of all 18 tracts in these groups).
We find that SGL and PCR-SGL select FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \includegraphics[width=0.33\textwidth]{als_gpca_coefs.pdf}
    % }
    % \vspace{-7.0cm}
    % \includegraphics[width=\textwidth]{als_gpca_results.pdf}
    {\phantomsubcaption\label{fig:class-results:probs}}
    {\phantomsubcaption\label{fig:class-results:coefs3d}}
    {\phantomsubcaption\label{fig:class-results:tract-profiles}}
    {\phantomsubcaption\label{fig:class-results:false-positives}}
    {\phantomsubcaption\label{fig:class-results:false-negatives}}
    \caption{
    {\bf PCR-SGL accurately and interpretably predicts ALS diagnosis.}
    {\bf (a)} Classification probabilities for ALS diagnosis, with
    controls on the left, patients on the right, predicted controls in
    blue, and predicted patients in orange. That is, orange dots on the left
    represent false positives, while blue dots on the right represent
    false negatives. We achieve {\protect\alsAccuracy}\% accuracy with an
    ROC AUC of {\protect\alsRocAuc}.
    {\bf (b)} PCR-SGL coefficients are presented on the core fibers of major
    fiber bundles. They exhibit high group sparsity and are concentrated
    in the FA of the corticospinal tract (CST).
    The brain is oriented with the right hemisphere
    in the foreground and anterior to the right of the page. The CSTL, CSTR,
    callosum forceps anterior (CFA), left arcuate (ARCL), and right arcuate (ARCR)
    bundles are indicated for orientation.
    {\bf (c)} PCR-SGL identifies three portions of the CST as important,
    where $\hat{\beta}$ (dashed line, right axis) has large values. These
    are centered around nodes 30, 65, and 90, corresponding to locations
    of substantial differences in FA between the ALS and control groups
    (shaded areas indicates standard error of the mean).
    {\bf (d)} Bundle profiles for false positive classifications. Line
    colors correspond to the marker edge color in the top left plot.
    These individuals have reduced FA in the CST portions which SGL
    identified as important. Their misclassification is coherent with the
    feature importance and the group differences in FA.
    {\bf (e)} Individual bundle profiles for false negative
    classifications. These individuals have bundle profiles which
    oscillate between the group means.
    \label{fig:class-results}
    }
\end{figure}

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the tract profile
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets
see \nameref{S3_Fig} to \nameref{S8_Fig} for the full tract profile information in all tracts/datasets.

@@

In contrast to the ALS classification case, the selected $\alpha$ values
indicate high global sparsity over group sparsity, with $\alpha = $
{\whLRatio}, {\hbnLRatio}, and {\ccLRatio}, for the WH, HBN, and Cam-CAN
datasets, respectively. The model weights are distributed over many different
tracts and dMRI tissue properties (see
\cref{fig:age-results:wh-coefs3d,fig:age-results:hbn-coefs3d,fig:age-results:cc-coefs3d}
and supporing figures in \nameref{S3_Fig} to \nameref{S8_Fig}).
This demonstrates that SGL is not coerced to produce overly sparse results
when a more accurate model requires a dense selection of features.
Furthermore, inspecting the portions of bundles with larger coefficients in
\cref{fig:age-results:wh-profile,fig:age-results:hbn-profile,fig:age-results:cc-profile}
reveals that SGL selects informative regions where diffusion properties are
different between the age quintiles.

@@

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \hspace{0.775cm}
    %     \includegraphics[width=0.26\textwidth]{wh_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{hbn_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{camcan_coefs.pdf}
    % }
    % \vspace{-6.1cm}
    % \includegraphics[width=\textwidth]{regression_scatter.pdf}
    {\phantomsubcaption\label{fig:age-results:wh-scatter}}
    {\phantomsubcaption\label{fig:age-results:hbn-scatter}}
    {\phantomsubcaption\label{fig:age-results:cc-scatter}}
    {\phantomsubcaption\label{fig:age-results:wh-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:hbn-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:cc-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:wh-profile}}
    {\phantomsubcaption\label{fig:age-results:hbn-profile}}
    {\phantomsubcaption\label{fig:age-results:cc-profile}}
    \phantomcaption{%
    {\bf Predicting age with tractometry and SGL.}
    {\bf (top)} The predicted age vs. true age of each individual from the test
splits (i.e., when each subject's data was held out in fitting the model) for
the {\bf (a)} WH, {\bf (b)} HBN, and {\bf (c)} Cam-CAN datasets; an accurate
prediction falls close to the $y=x$ line (dashed). The mean absolute error (MAE)
and coefficient of determination $R^2$ are presented in the lower right of each
scatter plot. {\bf (middle)} Feature importance for predicting age from
tract profile in the {\bf (d)} WH, {\bf (e)} HBN, and
{\bf (f)} Cam-CAN datasets. The orientation of the brain is that same as in
\cref{fig:class-results:coefs3d}, however because the coefficients exhibit high
global sparsity (as opposed to group sparsity), we plot the mean of the absolute
value of $\hat{\beta}$ for each bundle on the core fiber. The global
distrubution of the $\hat{\beta}$ coefficients reflects the fact that aging is
not confined to a single white matter bundle. {\bf (bottom)} Age quintile bundle
profiles for the {\bf (g)} WH, {\bf (h)} HBN, and {\bf (i)} Cam-CAN datasets.
    \label{fig:age-results}
    }
\end{figure}

\begin{figure}[b!]
    % \includegraphics[width=\textwidth]{model_performance.pdf}
    \caption{%
    {\bf Model performance across all datasets.}
Each panel shows model performance measured on the test set for each
cross-validation split, with each black dot representing a split, box
plots representing the quartiles, and white diamonds representing the mean
performance. The $y$-scale varies in each subplot.
\textbf{(a)} Accuracy of test set predictions for the
ALS dataset. Because group differences
in ALS diagnosis are mostly confined to a single bundle, the group
structure-preserving methods, SGL and PCR-SGL, outperform the other
models.
The remaining frames show coefficient of determination, \protect$R^2$ in test sets
for the \textbf{(b)} WH, \textbf{(c)} HBN, and \textbf{(c)} Cam-CAN
datasets. Because aging affects the white matter globally, group structure-blind
methods like elastic net and PCR Lasso perform well. Nonetheless, the SGL
models show competitve predictive performance, adapting to a problem where
group structure is not as informative. PCR-SGL performs poorly in this regime
because its initial group-wise PC projection destroys between bundle covariance.
The bundle-mean lasso performs poorly, demonstrating the value of along-tract profiling.
    \label{fig:model-performance}
    }
\end{figure}

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in \nameref{S3_Fig} to \nameref{S8_Fig}. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \cite{pinto2020harmonization}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.

@@

\item Diffusion MRI from a previous study of the corticospinal
tract (CST) in patients with amyotrophic lateral sclerosis
(ALS \cite{sarica2017corticospinal}), containing data from 24 ALS
patients and 24 demographically matched healthy controls. These data
were measured in a GE Discovery 750 3T MRI scanner at the Institute
of Bioimaging and Molecular Physiology in Catanzaro. Informed consent
was provided as approved by the Ethical Committee of the University
``Magna Graecia'' of Catanzaro. Voxel resolution was
$2 \times 2 \times 2 \text{mm}^3$ and 27 non-colinear directions were measured with a
$b=1000$ $\text{s} / \text{mm}^2$. Data was preprocessed to
correct for subject motion and for eddy currents. The diffusion tensor
model \cite{basser1994mr} was fit in every voxel.
We will refer to this dataset as ALS.

\item Diffusion MRI data from a previous study of properties of
the white matter across the lifespan \cite{yeatman2014lifespan},
containing dMRI data from 76 subjects with ages 6-50. These data were
measured in a GE Discovery 750 3T MRI scanner at the Stanford Center
for Cognitive and Neurobiological Imaging. The Stanford University
IRB approved the procedures of this study. Informed consent was
obtained from each adult participant, and assent for participation
was provided by parents/guardians for children. Voxel resolution was
$2 \times 2 \times 2 \textrm{mm}^3$ with 96 non-colinear directions measured with a
$b=2000$ $\textrm{s} / \textrm{mm}^2$ and 30 non-colinear directions
measured with a $b=1000$ $\textrm{s} / \textrm{mm}^2$. These data
were acquired using a twice refocused spin echo sequence, in which there is
sufficient time for eddy currents to subside between the application of
the gradients and the image acquisition, so no eddy current correction
was applied, but motion correction was applied before fitting the
diffusion tensor model \cite{basser1994mr} in every voxel using a robust
fit \cite{chang2005restore}
on the \protect$b=1000$ \protect$\textrm{s} / \textrm{mm}^2$ shell only.
We will refer to this dataset as WH.

\item Diffusion MRI data from the Healthy Brain Network pediatric mental
health study \cite{alexander2017open}, containing dMRI data from 1651 subjects
with ages 5-21. These data were measured in 3T Siemens MRI scanners at two
sites in the New York
area: Rutgers University Brain Imaging Center and the CitiGroup Cornell Brain Imaging Center. Age distributions of participants in the different sites are in \nameref{S9_Fig}.
Informed consent was obtained from each participant aged 18 or older. For
participants younger than 18, written consent was obtained from their legal
guardians and written assent was obtained from the participant. Voxel
resolution was $1.8 \times 1.8 \times 1.8 \text{mm}^3$ with 64 non-colinear directions
measured for each of $b=1000 \text{s} / \text{mm}^2$ and $b=2000 \text{s} /
\text{mm}^2$. Preprocessing was performed using \emph{QSIPrep} 0.12.1, which
is based on \emph{Nipype} 1.5.1 \cite[RRID:SCR\_002502]{nipype1,nipype2}.

@@

\item Diffusion MRI data from the Cambridge Centre for Ageing and
Neuroscience (Cam-CAN) ``CC700'' dataset
\cite{shafto2014cambridge,taylor2017cambridge}, containing data from 640
subjects with ages 18-88. These data were measured on a 3T Siemens TIM Trio
system and written informed consent was obtained from each participant. Voxel
resolution was $2 \times 2 \times 2 \text{mm}^3$ with 30 non-colinear directions
measured for each of $b=1000$ $\text{s} / \text{mm}^2$ and $b=2000$
$\text{s} / \text{mm}^2$. The diffusion weighted images were acquired
with a twice refocused spin-echo sequence and the same preprocessing
and reconstruction pipelines
used for the HBN dataset was applied to this data. We will refer to this
dataset as Cam-CAN.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox
(\texttt{mAFQ}, version 1.1 for WH and version 1.2 for ALS)
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Eighteen major
tracts, which are enumerated in \nameref{S1_Fig}, were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalanobis distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in \nameref{S1_Fig}.
See Fig S3 of \cite{kruper2021evaluating} for a depiction of these white matter bundles.
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (see \cref{fig:methods:group-structure}).

@@

For the multisite HBN study, we use the ComBat harmonization method to robustly adjust for site effects in the tract profiles. Initially designed to correct for site effects in gene expression studies \cite{johnson2007adjusting}, ComBat employs a parametric empirical Bayes approach to adjust for batch effects and has since been applied to multi-site cortical thickness measurements \cite{fortin2018harmonization}, multi-site DTI studies \cite{fortin2017harmonization}, and functional MRI data in the Adolescent Brain Cognitive Development Study (ABCD) \cite{nielson2018detecting}. We rely on the \texttt{neurocombat\_sklearn} library \cite{neurocombat_sklearn}, to apply ComBat in our scikit-learn analysis pipeline and present bundle profile site differences and ComBat correction in \nameref{S10_Fig} to \nameref{S13_Fig}.

@@

\begin{figure}[h!]
    % \includegraphics[width=0.75\textwidth]{nested-cross-validation.pdf}
    \phantomcaption{
    {\bf Nested cross-validation.}
We evaluate model quality using a nested $k$-fold cross validation
scheme. At level-0, the input data is decomposed into $k_0$ shuffled
groups and optimal hyperparameters are found for the level-0 training
set. To avoid overfitting, the optimal hyperparameters are themselves
evaluated using a cross-validation scheme taking place at level-1 of
the decomposition, where each level-0 training set is further
decomposed into $k_1 = 3$ shuffled groups. In the classification
case, the training and test splits are stratified by diagnosis. For
the ALS and WH data, $k_0 = 10$, while for the HBN and Cam-CAN data,
$k_0 = 5$.
    \label{fig:nested-cross-val}
    }
\end{figure}

@@

We would like to thank Scott Murray for a useful discussion of the SGL method
and Mareike Grotheer and John Kruper
for helpful comments on the manuscript.
Data collection and sharing for this project was provided by the Cambridge Centre for Ageing and Neuroscience (CamCAN). CamCAN funding was provided by the UK Biotechnology and Biological Sciences Research Council (grant number BB/H008217/1), together with support from the UK Medical Research Council and University of Cambridge, UK. This manuscript was prepared using a limited access dataset obtained from the Child Mind Institute Biobank, The Healthy Brain Network dataset. This manuscript reflects the views of the authors and does not necessarily reflect the opinions or views of the Child Mind Institute.

% \bibliography{paper}
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for
% step-by-step instructions.

\begin{thebibliography}{100}
